### 2012-08-02 paper/chemokine_main.tex
Paper split into Main and Supplement

During the primary immune response clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci representing a tiny fraction of the total lung.  T-cell navigation through the complex branching bronchial vascular network is made possible by local cytokine and chemokine production from infected epithelial cells.  Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.  Of the eight chemokines tested only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (lab results only) with IP-10 always dominating the migration direction.  A delayed differential equation model was fit to the empirical data revealing per cell chemokine production rates for each strain.  These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects inter-strain variation of chemokine production on T-cell recruitment.  The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.  Expanding plaque sizes further isolate infected cells, making T-cell discovery and subsequent clearance of the infection less effective over time.  The immune response is unable to clear the infection of pandemic H1N1 influenza due to the high rate of viral production and plaque size expansion.

@@

The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.  Understanding all components of defense mechanisms employed will lead to improved vaccine design and strategies to control immunopathology. Elegant studies in the mouse model have described the induction and effector phases of the adaptive immune response, including the functional capabilities of the innate interferon-directed response, macrophages, specific antibody, and cytotoxic effector lymphocytes \cite{Sallusto2000, Joo2008, Mackay2008}.   CD8+ T cells are necessary for the resolution of pneumonia and the complete clearance of mouse-adapted influenza strains \cite{Sallusto2000, Joo2008, Mackay2008, Miao2010}, but gaps remain in our understanding of the behavior of activated T cells in infected tissue.

The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naïve T and B cell precursors are encountered and activated in specialized architecture \cite{Saenz2010, Beltman2007, Handel2008, Zheng2008, Ingulli2009, Allan1990}.  T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneous in the lung, spleen and other organs 4-5 days post-infection after to extravasation from the blood, a whole-body distribution known as the lymphocyte diaspora \cite{Thelen2008}.  In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.  Once in the tissue, T cells are guided up a chemokine gradient to the infected epithelial cell secreting the chemokine.

A number of mathematical models have used detailed mouse model data to abstract the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.  Such whole-response modeling has considerable potential in selecting the most promising strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.  Models using ordinary differential equations to predict temporal events, however, can not account for the spatial constraints on cell behavior, and therefore make assumptions about the efficiency of each functional component in the model.

Here we focus on the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).  To develop a model with tractable simulations the model does not incorporate some key components of host defense and therefore does not aim to predict the true outcome for each strain examined.   We ask how do very small foci of infected tissue attract limited numbers of activated CD8 T cells after release into the systemic vascular system.  The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.   A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.

To provide data on the local chemokine-secretion environment in which T cells migrate, we used our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.   We selected in vitro chemokine data since this tissue environment surrogate may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and also permits the creation of sustained chemokine gradients within the model lung.  Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to populate the ABM model with activated CD8+ T cells.

The standard hypothesis would posit that in this model the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.  While we found this to be true, we observed in this spatially explicit model three interesting characteristics of chemokine-directed migration not revealed from experimental data.  First, of the only two chemokines stimulated by all 3 strains, the influence of IP-10 consistently dominated over the effect of RANTES, regardless of strain.  Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at higher concentrations above the threshold.  Finally, when the search space (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient and may be reflected in the loss of control of the pandemic strain.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.6\textwidth]{SystemChart}
\end{center}
\caption{A region of infected tissue of radius $r$ (shaded region $A$) expressing chemokines and inflammatory signals. This region is surrounded by region $B$ which does not have any infected cells and hence does not express either chemokines or inflammatory signals. The $2^{14}$ capillaries (red circles) are distributed evenly throughout the entire lung (region $A$ and $B$).}
\label{fig:systemchart}
\end{figure}

Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response. A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig 1) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions. To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T-cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T-cells that do not encounter cytokine recirculate to the lymph node. Once at the site of infection, a T-Cell may encounter an infected epithelial cell and induce apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine. Virus diffuses locally, infecting nearby cells, and continuing the cycle. The chemokine diffuses from secreting cells, thus increasing the size of the signaling area. After a delay to simulate lymph node stimulation and T-Cell proliferation, activated T-cells begin exiting the lymph node and circulating through the vasculature to the tissue. Because T-Cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations. Figure \ref{fig:modelchart} illustrates the model components 
(epithelial cells, T-cells, and particles) and how they transition between different states.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{ModelChart}
\end{center}
\caption{A visual representation of the model.  Healthy epithelial cells are infected by virus.  These infected cells secrete more virus after an incubation delay.  T-cells are activated in the lymph node and traverse the bronchial vascular network to the site of infection where they are recruited by inflammatory cytokine.  Once inside the lung tissue, T-cells climb the chemokine gradient and induce apoptosis in infected cells.}
\label{fig:modelchart}
\end{figure}

Epithelial cells are stationary and can be in of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}. \emph{healthy} cells remain unchanged unless infected by virus. Once infected, there is a 10h incubation window before the cell transitions from {incubating} to \emph{expressing}. \emph{Expressing} cells secrete virus and chemokine for 17 hours and then die. \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells. Apoptoic cells continue to secrete virus and die one hour after their transition. \emph{Dead} cells remain inert and do not regenerate over the course of an infection.

T-cells have three states: \emph{searching}, \emph{recirculating}, and \emph{activated}. T-Cells begin emerging from the lymph node at five days post infection at the rate of 777 T-cells per hour for the duration of the simulation. T-Cells enter the vascular system in the \emph{searching} state, arriving at a random location on the lung's surface. \emph{searching} cells perform a random walk in the tissue for 10 minutes, after which they transition to \emph{circulating} if they fail to encounter chemokine. \emph{Circulating} cells spend six minutes recirculating to the lymph node where they are converted to \emph{searching} and reintroduced to a new random location in the lung. If a \emph{searching} T-Cell encounters chemokine, it immedaiately converts to \emph{activated} and begins climbing the chemotactic gradient to the source of infection. \emph{Activated} T-Cells move continuously up the gradient and if they encounter \emph{expressing} epithelial cells, the epithelial cell transitions \emph{apotosis}. Activated T-cells are removed from the simulation using an exponential decay rate with an average lifespan of two hours.

The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \emph{expressing} epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration. Chemokine diffuses across the tissue but has no
direct effect beyond activating T-Cells. Both particle types decay over time.

\subsection*{Parameters}

Define CyCells environment.  Talk about the need for spatial model.  Human/mouse scaling.  Parameter definitions.

\begin{table}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
  Viral Decay &  $1/day$ & Phenomenological \\
  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
  Chemokine Decay &  $3.8508e-4/s$ & 30m half-life \\
  Infection Sensitivity &  $2 h/virion$ & Phenomenological \\
  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
  Apoptosis Time & $1 h$ & Phenomenological \\
  T-Cell Production Rate & $777/h$ & \cite{Miao2010} \\ 
  Circulation Time & $6 m$ & \cite{Banerjee2010b} \\
  Search Time & $10 m$ & \cite{Banerjee2010b} \\
  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Speed (Activated) & $3 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
  T-Cell Expected Kill Rate & $10 m$ & Fred/Drew \\
  Cell Radius & $25 \mu m$ & Phenomenological \\
  T-Cell Age (at FOI) & $2 h$ & [REF] \\
  T-Cell Age (Other) & $3 d$ & [REF] \\
  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
  IgM Strength & Viral decay of $3/day$ & Phenomenological \\
  \hline  
\end{tabular}
\caption{Generic model parameters}
\label{table:parameters}
\end{table}

@@

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either a seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or an avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 µl aliquots for a panel of 17 chemokines and cytokines (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.

@@

Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).  IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ODE model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells.  Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.  The model estimates are shown in Table~\ref{table:strains}.  These expression rates are similar across different strains with the exception of significantly higher RANTES production in aH5N1.  There is no discernible correlation between viral production and induced chemokine production across the three strains of influenza.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{data}
 \end{center}
\caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).  For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.  sH1N1 IP-10 registered three values above the measurement threshold of 8,500 $pg/mL$ (dotted line).  These values were not included in model fitting.} 
 \label{fig:data}
\end{figure}

\subsection*{Stochastic modeling effects}

Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure~\ref{fig:variance}).  To test the strength of this effect, we ran each model fifty times using the default parameters given in Table~\ref{table:parameters}.  $R^2$ values were calculated for individual runs versus the average of all runs.  Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9935 and standard deviation of 0.0029.  sH1N1 had a mean $R^2$ 0.9859 with a standard deviation of 0.0109.  aH5N1 had a mean $R^2$ of 0.9167 with a standard deviation of 0.0726.

Each run took the calculated viral production and chemokine production rates for the three different strains of influenza as input and reported the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.  Therefore the figures approximate plaque growth over time.

Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the presence of IgM.  In addition, for each infection the number of infected cells declines quickly at day five due to the T-cell response.

The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).  The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.  By the time T-Cells arrive, the plaque is too large for them to cover effectively.  Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T-Cells to gain ground is too small.  aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{variance}
 \end{center}
\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray). IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.  Each run was initialized identically for each strain.  The middle line shows the average while the outer lines show the 95\% confidence interval.} 
 \label{fig:variance}
\end{figure}

@@

T-Cells exit a capillary when they sense inflammatory cytokine and then follow the chemokine gradient in tissue.  We use the chemokine gradient as a proxy for the inflammatory response.  Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.  T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.

To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude. We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 100 and 10 $ng/ml$, below which there is no discernible effect on infection kinetics.  Because multiple T-cells in the same area do not increase the rate at which apoptosis is induced, we hypothesize that a critical number of T-cells are required to clear an infection above which there is no additional benefit gained from increasing T-cell presence.  Reference \cite{Gao2003} reports T-cell chemotaxis at concentrations as low as 1 $nM$ ($10 ng/mL$ assuming a chemokine weight of $10 kDa$), therefore we use that sensitivity for each model run.

@@

\subsection*{Chemokine combinations}

Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 ineffectual.  We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table~\ref{table:strains}).  Because of this behavior IP-10 was not included in the aH5N1 model runs.

Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~\ref{fig:chemokine}).  The lack of both chemokines leads to runaway infections in all three strains.  The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.  IP-10 alone proves to be as effective as the combination of IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.

\begin{figure}[ht!]
\begin{center}
	\includegraphics[width=\textwidth]{chemokine}
	\caption{Effects of different chemokine combinations.  A) aH5N1 does not stimulate an IP-10 response.  B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
	\label{fig:chemokine}
\end{center}
\end{figure}

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spread and clearance of infection.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T-cells to the infected cells.  Although virus is produced at a higher rate than chemokine, it diffuses much more slowly because it is a larger particle.  Chemokine diffusion, however, is limited by higher decay rates than the virus.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).

>>>>

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci representing a tiny fraction of the total lung.  T-cell navigation through the complex branching bronchial vascular network is enabled by local cytokine and chemokine production from infected epithelial cells.  Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.  Of the eight chemokines tested only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (results only) with IP-10 always dominating the affected T-cell migration direction.  A delayed differential equation model was fit to the empirical data, revealing per cell chemokine production rates for each strain.  These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects of inter-strain variation of chemokine production on T-cell recruitment.  In the model immune response is unable to clear the infection of pandemic H1N1 influenza due to the high rate of viral production and plaque size expansion.  The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.  Expanding plaque sizes isolate infected cells, causing T-cell discovery and subsequent clearance of the infection to become less effective over time.

@@

The adaptive immune response induced during acute influenza A infection is a complex web of defense mechanisms that controls all but the most virulent strains.  Understanding its components and their interactions are relevant to improving vaccines and strategies to control immunopathology. Elegant studies in the mouse model have described the induction and effector phases of the adaptive immune response, including functional capabilities of the innate interferon-directed response, macrophages, specific antibody, and cytotoxic effector lymphocytes \cite{Sallusto2000, Joo2008, Mackay2008}.   CD8+ T cells are necessary for the resolution of pneumonia and complete clearance of mouse-adapted influenza strains \cite{Sallusto2000, Joo2008, Mackay2008, Miao2010}, but gaps remain in our understanding of the behavior of activated T cells in infected tissue.

The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naive T and B cell precursors are activated in specialized architecture \cite{Saenz2010, Beltman2007, Handel2008, Zheng2008, Ingulli2009, Allan1990}.  T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneously in the lung, spleen and other organs 4-5 days post-infection, a whole-body distribution known as the lymphocyte diaspora \cite{Thelen2008}.  In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.  Once in the tissue, T cells follow a chemokine gradient to the infected epithelial cells secreting the chemokine.

Mathematical models have used detailed mouse data to characterize the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.  Such whole-response modeling has considerable potential to inform strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.  Models using ordinary differential equations to predict temporal events, however, do not account for spatial constraints, and must make assumptions about the efficiency of the spread and interaction of populations.

Here we focus on the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).  To develop a model with tractable simulations, some key components of host defense are not incorporated explicitly and the model therefore does not aim to predict the true outcome for each strain examined.   We ask how do very small foci of infected tissue attract limited numbers of activated CD8 T cells after release into the systemic vascular system.  The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.   A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.

To characterize the local chemokine-secretion environment in which T cells migrate, we used data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.   We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients within the model lung.  Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to populate the ABM model with activated CD8+ T cells. [REF]

The standard hypothesis would posit that in this model the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.  While we found this to be true, we observed three interesting characteristics of chemokine-directed migration not revealed from experimental data.  First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.  Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at increasing concentrations above the threshold.  Finally, when the target (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient, possibly contributing to the loss of control of the pandemic strain.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.5\textwidth]{SystemChart}
\end{center}
\caption{A simplified model of T cell search.  Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through the 14 layers of the branching bronchial network.  Upon reaching a capillary, T cells will exit into the tissue in the presence of a cytokine signal.  T cells that do not sense a signal will continue to recirculate either through the lymph network or through the pulmonary vein back to the top of the network.}
\label{fig:systemchart}
\end{figure}

Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions. To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node. At the site of infection when a T cell encounters an infected epithelial cell it induces apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine. Virus diffuses locally, infecting nearby cells, and continuing the cycle. Chemokine also diffuses from secreting cells, creating a ball of stimulation around the infected region. After a delay to simulate lymph node stimulation and T cell proliferation, activated T cells begin exiting the lymph node and circulating through the vasculature to the tissue. Because T  cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{ModelChart}
\end{center}
\caption{A visual representation of the model.  Healthy epithelial cells infected by virus begin secreting virus after an incubation delay.  T cells are activated in the lymph node and traverse the bronchial vascular network to the site of infection where they are recruited by inflammatory cytokine.  Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.  Solid arrows represent the transition of a cell from one behavioral state to another.  Dashed arrows display the mechanism used to induce a transition.  Dotted arrows show the production of new virus.}
\label{fig:modelchart}
\end{figure}

Epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}. \emph{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from {incubating} to \emph{expressing}. \emph{Expressing} cells secrete virus and chemokine for 17 hours and then die. \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells. Apoptoic cells continue to secrete virus until they die one hour after their transition. \emph{Dead} cells remain inert and do not regenerate over the course of an infection.

T-cells have three states: \emph{searching}, \emph{recirculating}, and \emph{chemotaxing}. \emph{Searching} T-Cells begin emerging from the lymph node at five days post infection. These T-Cells arrive at a random location on the lung's surface, perform a random walk in the tissue for 10 minutes and transition to \emph{circulating} in the absence of chemokine. \emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{searching} and are reintroduced to a new random location in the lung. When a \emph{searching} T-Cell encounters chemokine, it converts to \emph{chemotaxing} and begins climbing the chemotactic gradient to the source of infection. \emph{Chemotaxing} T-Cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells. \emph{Chemotaxing} decay exponentially with an average lifespan of two hours.

The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \emph{expressing} epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration. Chemokine diffuses across the tissue but has no
direct effect beyond activating T-Cells. Both particle types decay exponentially.

Parameters that are consistent between every model are shown in Table \ref{table:parameters}.  Strain-specific values are shown in Table \ref{table:strains}.

\begin{table}
\begin{center}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
  Viral Decay &  $1/day$ & Phenomenological \\
  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
  Chemokine Decay &  $3.8508$e-4$/s$ & 30m half-life \\
  Infection Sensitivity &  $2 h/virion$ & Phenomenological \\
  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
  Apoptosis Time & $1 h$ & Phenomenological \\
  T-Cell Production Rate & $777/h$ & \cite{Miao2010} \\ 
  Circulation Time & $6 m$ & \cite{Banerjee2010b} \\
  Search Time & $10 m$ & \cite{Banerjee2010b} \\
  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Speed (Activated) & $3 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
  T-Cell Expected Kill Rate & $10 m$ & [REF] \\
  Cell Radius & $25 \mu m$ & Phenomenological \\
  T-Cell Age (at FOI) & $2 h$ & [REF] \\
  T-Cell Age (Other) & $3 d$ & [REF] \\
  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
  IgM Strength & Viral decay of $3/day$ & Phenomenological \\
  \hline  
\end{tabular}
\caption{Generic model parameters}
\label{table:parameters}
\end{center}
\end{table}

@@

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 µl aliquots for a panel of 17 chemokines and cytokines (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.

@@

Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).  IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ODE model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells.  Model fits, shown in Table \ref{table:strains}, were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.  These expression rates are similar across different strains with the exception of significantly higher RANTES production in aH5N1.  There is no discernible correlation between viral production and induced chemokine production across the three strains.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{data}
 \end{center}
\caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).  For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.  sH1N1 IP-10 registered three values not included in model fitting above the measurement threshold of 8,500 $pg/mL$ (dotted line).} 
 \label{fig:data}
\end{figure}

@@

T-Cells exit a capillary when they sense inflammatory cytokine and follow the chemokine gradient.  The model uses chemokine gradient as a surrogate for the inflammatory response.  Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.  T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.

To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude. We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on infection kinetics.  Because multiple T-cells in the same area do not increase apoptosis, we hypothesize that above a critical threshold no additional benefit is conferred by increasing T cell presence.  We use the sensitivity of 10 $ng/mL$ for each run, which corresponds to a $1 nM$ concentration assuming a chemokine molecular weight of 10 $kDa$ \cite{Gao2003}.

@@

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T-cells to the infected cells.  Although virus is produced at a higher rate than chemokine, the larger particle diffuses much more slowly.  Chemokine diffusion decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).

Up to day 4 the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare. Over time, cells in the plaque's interior die, and active cells are a decreasing proportion of the plaque. T-cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  In H5N1, the plaque is dense, allowing T-cells to find secreting cells easily and infection is eliminated.  However, in both H1N1 simulations secreting cells were not eliminated.  Expressing cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T-cells can lag behind cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T-cells from completely clearing the infection.

The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig. \ref{fig:sensitivity}).  T-cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}). By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T-cells can control (but not eliminate) the sH1N1 infection.

A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.  That is, a new viral particle is produced approximately every 200 seconds.  Assuming that secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit production to 18 new viral particles.  T-Cells alone cannot contain the pH1N1 infection.  In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T-cells to limit a single infected cell to 1.3 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.
